---
title: "Analysis of Hospital Data"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r}

if (!require("readxl")){ install.packages("readxl")}
if (!require("pander")){ install.packages("pander")}
if (!require("summarytools")){ install.packages("summarytools")}
if (!require("psych")){ install.packages("psych")}
if (!require("tidyverse")){ install.packages("tidyverse")}
if (!require("epitools")){install.packages("epitools")}
if (!require("gmodels")){install.packages("gmodels")}

library(readxl)
library(tidyverse)
library(pander)
library(summarytools)
library(psych)
library(epitools)
library(gmodels)

```

Getting the data into the session from excel. 

```{r}
hospital <- read_excel("hospital_data_2021.xls")

glimpse(hospital)

```

Here's what variables are in the  *hospital* dataset: 

- *age* (years)
- *bp* blood pressure
- *lengthstay* length of individuals stay (days)
- *ethnic* Ethnicity (in 7 categories)
- *sex* 
- *day* (day of the week admitted)
- *pulse* resting pulse rate
- *prevhospad* (previous hospital admissions) 
- *charlson* - a marker of co-morbidity
- *agegroup* 
- *dead* (did they die during their hospital stay)

We are going to now do just a little bit of tidying up of the data  

We need to tell R which of these are categorical (factors) and what labels to give for the factors. Run the R command below to do this. 

```{r}
hospital$ethnic <- factor(hospital$ethnic, levels=c(1,2,3,4,5,6,7,8,9,10), labels = c("white", "black carribean", "black african","black other","indian","pakistani", "bangladeshi", "chinese", "other", "missing"))

hospital$sex <- factor(hospital$sex, levels=c(1,2), labels = c("male","female"))

hospital$day <- factor(hospital$day, levels=c(1,2,3,4,5,6,7), labels=c("Friday","Monday","Saturday","Sunday","Thursday","Tuesday","Wednesday"))
                                                                       
                                                                  
hospital$dead <- factor(hospital$dead, levels=c(0,1), labels = c("alive","dead"))

hospital$agegroup <- factor(hospital$agegroup, levels=c(1,2,3,4,5,6), labels=c("under 50","50-59", "60-69", "70-79", "80-89","90 and above"))
```

```{r}
summary(hospital)

hist(hospital$age)
```



1)	Is there a difference in blood pressure *bp* at admission (outcome) between those who die and those who do not die *dead* (exposure)? 

First plot the histogram- a reminder of the code generically hist(dataset$outcome)
```{r}
hist(hospital$bp)
```


Now conducted the t-test for the difference in blood pressure between those who died and did not die in hospital.

Generically:

t.test(dataset$outcomevariable ~ datset$exposurevariable)


```{r}
t.test(hospital$bp ~ hospital$dead)
```

What is the mean  blood pressure in each group?

Dead  = 71.7
Alive = 76.9

Mean Difference Alive-dead (you need to do this by hand) = 5.2

What is the 95% CI around the mean difference? 95% CI = 1.66 - 8.77

What is the p=value? = 0.00441

Is the p-value significant? Yes, assuming your threshold was set at p <= 0.05 

Describe what this p-value means. 



2)	Now have a look at whether pulse *pulse* rate differs between those who lived and those who died *dead*. 

First plot the histogram
```{r}
hist(hospital$pulse)
```


Now conducted the t-test for the difference in pulse between those who died and did not die in hospital.

```{r}
t.test(hospital$pulse ~ hospital$dead)

```

What is the mean pulse rate in each group?

Dead  = 92.8
Alive = 88.4

Mean Difference Alive-dead  = 4.4

What is the 95% CI around the mean difference?  95% CI = -9.6 - 0.75

What is the p=value = 0.09286



**Chi-square tests and Odds Ratios**

Investigate the association between sex *sex*  and death *dead* (the variable indicating whether people die during the admission). Conduct a chi-squared test and interpret the p value. 


CrossTable(datset$exposurevariable,  dataset$outcomevariable, chisq=TRUE)

The CrossTable command gives you a cross tabulation of two variables. It shows the following in the different row:
row 1 = N - observed number 
row 2 = chi-square contribution
row 3 = row percentages  (this will be the row we look at to get the prevalence for males and females)
row 4 = column percentages
row 5 = table total %- not really used

```{r}
CrossTable(hospital$sex, hospital$dead, chisq = TRUE)
```


What percentages of males  died?  

What percentages of females died?

Look at the p-value from the Pearson's Chi-square test (use this for the p-value unless your dataset is small).  What is this value? p = 0.439

Is this a significant association? No



Now use R to calculate the odds ratios using the command below:

Generically:
epitab(dataset$exposurevariable, dataset$outcomevariable)

```{r}
epitab(hospital$sex, hospital$dead)
```

Describe the odds ratio for females to males and provide the 95% CI:

+type answer here

  


Now examine the association between previous hospital admission in the last year  *prevhospad* and death *death*.

Chi-square test
```{r}
CrossTable(hospital$prevhospad, hospital$dead, chisq = TRUE)
```

What percentages of people who were not previous hospitalised (coded 0) died?

What percentages of people who were previously  hospitalised (coded as 1) died?

Look at the p-value from the Pearson's Chi-square test.  What is this value?  p=

Is this a significant association?



Odds ratio for the association between previous hospital admission *prevhospad* and death *death*

```{r}
epitab(hospital$prevhospad, hospital$dead)
```

Describe the odds ratio for an hospital admission in the previous year and odds of death and the describe the 95% CI

+type answer here




The association between day of the week of admission *day* and death *dead*. Conduct a chi-squared test and interpret the p value.

```{r}
CrossTable(hospital$day, hospital$dead, chisq = TRUE)
```


What day of the week did the greatest %  people die on? Sunday

What day of the week did the least % of people die on? Friday, Monday, Tuesday

Look at the p-value from the Pearson's Chi-square test.  What is this value? p = 0.424

Is this a significant association?- No


Describe the odds ratio for the highest level of co-morbidity to the lowest level.

+type answer here
```{r}
head(hospital$lengthstay)
summary(hospital$lengthstay)
```
```{r}
##Logistic regression 
hospital_coded <- read_excel("hospital_data_2021.xls")

hospital_coded$ethnic <- factor(hospital_coded$ethnic, levels=c(1,2,3,4,5,6,7,8,9,10), labels = c("white", "black carribean", "black african","black other","indian","pakistani", "bangladeshi", "chinese", "other", "missing"))

hospital_coded$sex <- factor(hospital_coded$sex, levels=c(1,2), labels = c("male","female"))

hospital_coded$day <- factor(hospital_coded$day, levels=c(1,2,3,4,5,6,7), labels=c("Friday","Monday","Saturday","Sunday","Thursday","Tuesday","Wednesday"))
                                                                       

hospital_coded$agegroup <- factor(hospital_coded$agegroup, levels=c(1,2,3,4,5,6), labels=c("under 50","50-59", "60-69", "70-79", "80-89","90 and above"))

#Fit the model
glm_fit <- glm(hospital_coded$dead ~ hospital_coded$age + hospital_coded$bp, + hospital_coded$lengthstay + hospital_coded$lengthstay, data = hospital_coded, family = binomial) 
summary(glm_fit)
```

```{r}
library(caret)
glm.probs <- predict(glm_fit,type = "response")
glm.probs[1:5]
```
 
 
```{r}
glm.pred <- ifelse(glm.probs > 0.5, "Up", "Down")
```
 
```{r}
attach(hospital_coded)
table(glm.pred, dead)
```
 
 
 